### 2017-11-10 vipassana-for-hackers.tex
done up to blue bars 1

\usepackage{hyperref}
\usepackage{graphicx}

@@

Put more succinctly, your activities will include: Eat, sleep, poop, bath, meditate.

That's it.

Meditation will consume approximately 10 to 11 hours per day.

@@

The following section will require some visual assistance, for which I will use the same slides I used in the talk which corresponds to this paper.

They look like this:

@@

Abrash discusses the nature of the ``Matrix Paradox'' \cite{simulationhypothesis} as simply the right neurons firing to simulate any particular environment. ``How do we know we are not in the matrix?'' The question is valid, but the diagram is missing a great deal of sensory input. The following might be a slightly better model of reality:

Though a classic sense diagram such as Mr. Abrash’s might describe proprioception \cite{ramachandranbrainfunction, proprioception}, nocioception \cite{nocioception}, and hunger \cite{fmrihunger} as physical ``feeling'', they are definitively not a subset of ``touch'', as they correspond to entirely different activity in the brain. Touch is strictly an external sense dealing with skin. These are internal senses.

@@

On that note, I feel I should introduce an absolutely fantastic RadioLab episode (``Where Am I?'' \cite{whereami}) which addresses this topic of ``other senses'' from the perspectives brought by modern Neuroscience. On the topic of lacking proprioception: ``We have words like `deaf' and `blind'. We do not have words for being deaf or blind to one's own body.'' In the Matrix, perhaps this is either a bug or a feature flag.

@@

On Attention

@@

Example 1:
A bustling restaurant

@@

Assuming we've just placed food in our mouths, our sense of taste is going off like crazy, we're hearing a lot of scattered sound around us, and we're also sufficiently distracted by most of our other senses. Note that we are probably paying very little attention to the internal senses.
Example 2:
A sensory deprivation tank

The corresponding sense map is on the right.

@@

Example 3:
An anechoic chamber

The corresponding sense map is again on the right.

@@

The above three examples serve as templates with external input guiding one's understanding of what is happening neurologically.Anapana Meditation:
Practicing Focus

The first 3.5 days of a 10-day Vipassana course are spend practicing ``Anapana'' (breath) meditation. This period is very important as it is time spent practicing acute, one-pointed attention. If we imagine our attention as a pink dot, the goal is the following clown effigy:

@@

On the first day, your attention might look something like this:
…where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent ``outside'' the body on thoughts and emotions. By the fourth day, your attention will often take a shape like this instead:

If we revisit the notion that ``meditation is a thing'', we can see a specific example here in Anapana:

As much as possible, we are forcing our attention to the sense of touch. In the above diagram, our attention is (realistically) not perfectly tuned and narrowed to a tiny spot under our nose. We are still likely to hear sounds around us on the 3rd day of the course, especially if the sound is monkeys clambering over the roof of the meditation hall.

Vipassana Meditation:
Mutually-Recursive
Consciousness I/O

Oddly enough, Vipassana mostly seems to be about remaining calm. If you're thinking ``you're sitting there, silent, for 10 hours a day... what is there to get upset about?'' you are in for a surprise. But first, a quick overview of the early mechanics of Vipassana:
Attention is moved from the nostril area to the top of the head, then systematically moved across every body part, down to the toes. This process then repeats. More complex variations are added as the bodily sensation experienced through this probing of the nervous system becomes increasingly subtle. In this order:

Add toes-to-head to head-to-toes
Dual/symmetrical observation (both arms, both legs)
Observation of a ``ring'' around the entire body
Observation of as much of the body as possible (while moving attention up and down)
Penetrating attention of muscles, tissues, organs
Penetrating attention of the spinal cord
Randomized attention in fingertip-sized dots throughout the body, inside and out
Sweeping attention through the entire body, up and down, in a single breath

Through this process, attention will move from the sense of touch to all the internal senses and the sense map will look more like this:

I should note that I have never successfully penetrated my muscles or inner tissues much. The average new meditator is likely to feel a great many fascinating sensations on her fingertips, palms, arms, toes, and legs. The torso, since it is surrounded in longer nerve fibres, will take longer to observe in the same way.

As a Canadian who spent little (if any) of my childhood sitting cross-legged on the floor, the following diagram roughly summarizes most of my meditation experiences:
...where the purple dots are intense pain and the pink dot continues to be my awareness. At the peak of my meditation practice toward the end of a course, I would find that if my attention were somewhere away (say, my neck) from the painful area of my body (say, my knees), I may not feel any pain at all. As I moved my attention around, the pain in my knees would come back as I approached them, but every time I could sustain focus on another part of my body, the pain would appear to decrease with every return visit to my knees. If I can remain calm and focused while away from my knees, my knees' pain seems to take care of itself. If I can remain calm and focused while my attention moves across my knees, the situation continues to improve.

If you can’t feel pain in your knees despite the fact that the pain is ``real'', you have effectively eliminated (some) sensory input. This is possible for all sensory input. The 5 classic sense-doors will quite clearly shut down in deep meditative states. Just because your eyes are closed does not make you blind. Light sensitivity remains, and as a side bonus of an optic nerve injury from a few years ago, I can ``see'' something (the damaged spot of the nerve) even when my eyes are shut and I'm in complete darkness. This blob which I ``see'' disappears in deep meditative states. Similarly, these deep states of meditation will eliminate audio input as well, which is easier for anyone to detect.
Blue Bars:
The New Distractions

@@

@@

>>>>

\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage{enumitem}
\setlist{noitemsep}

@@

Put more succinctly, your activities will include: Eat, sleep, poop, bath, meditate.\\

That's it.\\

Meditation will consume approximately ten to eleven hours per day. All courses the world over are free and volunteer-run. Material goods are provided for by donations from previous students.

@@

The following section will require some visual assistance, for which I will use the same slides I used in the talk which corresponds to this paper. An unmodified slide is seen here in Figure~\ref{fig:michael-abrash-senses}.

@@

Abrash discusses the nature of the ``Matrix Paradox'' (more commonly known as the ``Simulation Hypothesis'' \cite{simulationhypothesis}) as simply the right neurons firing to simulate any particular environment. ``How do we know we are not in the matrix?'' The question is valid, but the diagram is missing a great deal of sensory input. Figure~\ref{fig:michael-abrash-senses-extended} is a slightly better model of reality.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/michael-abrash-senses-extended-attached.jpeg}
  \caption{A more complete, 7-sense brain mapping.}
  \label{fig:michael-abrash-senses-extended}
\end{figure}

Though a classic sense diagram such as Mr. Abrash's might describe proprioception \cite{ramachandranbrainfunction, proprioception}, nocioception \cite{nocioception}, and hunger \cite{fmrihunger} as physical ``feeling'', they are definitively not a subset of ``touch'', as they correspond to entirely different activity in the brain. Touch is strictly an external sense dealing with skin. These are internal senses.

@@

On that note, I feel I should introduce an absolutely fantastic RadioLab episode (\textit{Where Am I?} \cite{whereami}) which addresses this topic of ``other senses'' from the perspectives brought by modern Neuroscience. On the topic of lacking proprioception: ``We have words like `deaf' and `blind'. We do not have words for being deaf or blind to one's own body.'' In the Matrix, perhaps this is either a bug or a feature flag.

@@

\begin{center}
  \Huge{On Attention}
\end{center}

@@

\begin{center}
  \LARGE{Example 1:}\\
  \LARGE{A bustling restaurant}
\end{center}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/attention1-restaurant.jpg}
  \caption{A bustling restaurant.}
  \label{fig:bustling-restaurant}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=10cm]{images/ma-noisy-restaurant.png}
  \caption{Sense map of a bustling restaurant.}
  \label{fig:sense-map-bustling-restaurant}
\end{figure}

Assuming we've just placed food in our mouths, our sense of taste is going off like crazy, we're hearing a lot of scattered sound around us, and we're also sufficiently distracted by most of our other senses. Note that we are probably paying very little attention to the internal senses.

\begin{center}
  \LARGE{Example 2:}\\
  \LARGE{A sensory deprivation tank}
\end{center}

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.48\linewidth}
    \includegraphics[width=\linewidth]{images/attention2-tank.jpg}
  \end{subfigure}
  \begin{subfigure}[b]{0.48\linewidth}
    \includegraphics[width=\linewidth]{images/ma-sensory-deprivation-tank.png}
  \end{subfigure}
  \caption{A sensory deprivation tank and corresponding sense map.}
  \label{fig:sensory-deprivation-tank}
\end{figure}

@@

\begin{center}
  \LARGE{Example 3:}\\
  \LARGE{An anechoic chamber}
\end{center}

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.48\linewidth}
    \includegraphics[width=\linewidth]{images/attention3-chamber.png}
  \end{subfigure}
  \begin{subfigure}[b]{0.48\linewidth}
    \includegraphics[width=\linewidth]{images/ma-anechoic-chamber.png}
  \end{subfigure}
  \caption{An anechoic chamber and corresponding sense map.}
  \label{fig:anechoic-chamber}
\end{figure}

@@

The above three examples serve as templates with external input guiding one's understanding of what is happening neurologically.

\begin{center}
  \Huge{Anapana Meditation:}\\
  \Huge{Practicing Focus}
\end{center}

The first 3.5 days of a 10-day Vipassana course are spend practicing ``Anapana'' (breath) meditation. This period is very important as it is time spent practicing acute, one-pointed attention. If we visualize our attention as any skin (internal or external; lips, nose, nostrils, etc.) covered by the pink triangle in Figure~\ref{fig:burmese-anapana-triangle}, we will get an idea of exactly where one's attention is directed at the beginning of Anapana practice.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/burmese-anapana-triangle.png}
  \caption{Large, triangular Anapana meditation target.}
  \label{fig:burmese-anapana-triangle}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/face-attention.jpg}
  \caption{Wandering attention.}
  \label{fig:wandering-attention}
\end{figure}

On the first day, your attention might look something like Figure~\ref{fig:wandering-attention}, where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent ``outside'' the body on thoughts and emotions. By the fourth day, your attention will often take a shape something like Figure~\ref{fig:narrow-attention}. Over the course of the first three days, the instructions narrow down one's attention from the nose-and-lips triangle from Figure~\ref{fig:burmese-anapana-triangle} to a finger-tip-sized area wherever the breath can be felt beneath the nostrils.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{images/face-attention-narrow.jpg}
  \caption{Narrow attention.}
  \label{fig:narrow-attention}
\end{figure}

If we revisit the notion that ``meditation is a thing'', we can see a specific example in Anapana with the help of the sense map in Figure~\ref{fig:anapana-sense-map}. Where mindfulness and violin, albeit useful physical and mental practices, are not meditation, we can see from the sense map that Anapana \textit{is} meditation. It is the practice of narrowing one's attention, repeatedly, toward a specific sense (in this case, touch) at a specific time (now) in a specific place (just beneath the nostrils). As our skill in Anapana improves, we come nearer and nearer to the focus-oriented meditation ideal of a single point of awareness with unwavering focus.

\begin{figure}[H]
  \centering
  \includegraphics[width=10cm]{images/ma-anapana.png}
  \caption{Anapana sense map.}
  \label{fig:anapana-sense-map}
\end{figure}

In Figure~\ref{fig:anapana-sense-map}, we can see our attention is (realistically) not perfectly tuned and narrowed to a tiny spot under our nose. We are still likely to hear sounds around us on the third day of the course, especially if the sound is monkeys clambering over the roof of the meditation hall.

\begin{center}
  \Huge{Vipassana Meditation:}\\
  \Huge{Mutually-Recursive}\\
  \Huge{Consciousness I/O}
\end{center}

Oddly enough, Vipassana mostly seems to be about remaining calm. If you're thinking ``you're sitting there, silent, for 10 hours a day... what is there to get upset about?'' you are in for a surprise. But first, a quick overview of the early mechanics of Vipassana. (See Figure~\ref{fig:vipassana-mechanics}.)
Attention is moved from the nostril area to the top of the head, then systematically moved across every body part, down to the toes. This process then repeats.

This visualization and description is to act as an \textit{example only}: this is not the only way to begin Vipassana meditation. Though Vipassana instruction is systematic, it is far from mechanical or prescriptive. Even in the initial instructions, alternative approaches are offered including alternate orders of observation, alternate areas of observation, and alternate durations of observation. The key aspect of any approach to Vipassana is that the \textbf{entire body} is repeatedly observed within each meditation period.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/burmese-vipassana-mechanics.png}
  \caption{Simplified example of initial Vipassana mechanics.}
  \label{fig:vipassana-mechanics}
\end{figure}

I will take this opportunity to echo a caveat from the beginning of this paper: This section is in no way a form of instruction. Do not attempt full-body Vipassana meditation outside of a proper 10-day course. Practicing Vipassana is likely the most difficult activity you will ever attempt, should you choose to. To try so alone, you will have no teacher for guidance and the experience is bound to be frustrating.

Consider this warning to also extend to any literature which \textbf{does} attempt to convey Vipassana meditation instruction. Specifically, Glickman's \textit{Beyond the Breath: Extraordinary Mindfulness Through Whole-Body Vipassana Meditation} \cite{glickman} includes a facsimile of instruction provided on a 10-day introductory Vipassana course. Instruction on a course includes 11+ hours of video instruction, at least twice as many hours of audio instruction, and multiple interviews per day with the teacher. Trying to replace this with a book is both inadequate and dangerous.\\

More complex variations are added as the bodily sensation experienced through this probing of the nervous system becomes increasingly subtle. In this order:

\begin{enumerate}
  \item Add toes-to-head to head-to-toes
  \item Dual/symmetrical observation (both arms, both legs)
  \item Observation of a ``ring'' around the entire body
  \item Observation of as much of the body as possible (while moving attention up and down)
  \item Penetrating attention of muscles, tissues, organs
  \item Penetrating attention of the spinal cord
  \item Randomized attention in fingertip-sized dots throughout the body, inside and out
  \item Sweeping attention through the entire body, up and down, in a single breath
\end{enumerate}

Through this process, attention will move from the sense of touch to all the internal senses and the sense map will look more like Figure~\ref{fig:vipassana-sense-map}.

\begin{figure}[h]
  \centering
  \includegraphics[width=8cm]{images/ma-vipassana1.png}
  \caption{Vipassana sense map.}
  \label{fig:vipassana-sense-map}
\end{figure}

I should note that I have never successfully penetrated my muscles or inner tissues much. The average new meditator is likely to feel a great many fascinating sensations on her fingertips, palms, arms, toes, and legs. The torso, since it is surrounded in longer nerve fibres, will take longer to observe in the same way. \textit{[citation needed]}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{images/burmese-vipassana-pain.png}
  \caption{Common pain for Westerners in meditative postures.}
  \label{fig:burmese-vipassana-pain}
\end{figure}

As a Canadian who spent little (if any) of my childhood sitting cross-legged on the floor, Figure~\ref{fig:burmese-vipassana-pain} roughly summarizes most of my early meditation experiences, where the purple dots are intense pain and the pink dot continues to be my awareness. At the peak of my meditation practice toward the end of a course, I would find that if my attention were somewhere away (say, my neck) from the painful area of my body (say, my knees), I may not feel any pain at all. As I moved my attention around, the pain in my knees would come back as I approached them, but every time I could sustain focus on another part of my body, the pain would appear to decrease with every return visit to my knees. If I can remain calm and focused while away from my knees, my knees' pain seems to take care of itself. If I can remain calm and focused while my attention moves across my knees, the situation continues to improve.

If an individual cannot feel pain in one's knees despite the fact that the pain is ``real'', she has effectively eliminated (some) sensory input. This is possible for all sensory input. The five classic sense-doors will quite clearly shut down in deep meditative states. Just because your eyes are closed does not make you blind. Light sensitivity remains, and as a side bonus of an optic nerve injury from a few years ago, I can ``see'' something (the damaged spot of the nerve) even when my eyes are shut and I'm in complete darkness. This blob which I ``see'' disappears in deep meditative states. Similarly, these deep states of meditation will eliminate audio input as well, which is easier for anyone to detect.

\begin{center}
  \Huge{Blue Bars:}\\
  \Huge{The New Distractions}
\end{center}

@@

\bibitem{fiveprecepts}
  Edited by Access to Insight.
  \url{https://www.accesstoinsight.org/ptf/dhamma/sila/pancasila.html}
  \textit{The Five Precepts: pañca-sila}
  Access to Insight (BCBS Edition), 30 November 2013.

@@

\bibitem{glickman}
  Marshall Glickman.
  \textit{Beyond the Breath: Extraordinary Mindfulness Through Whole-Body Vipassana Meditation}.
  Tuttle Publishing, 2002.
  ISBN: 1582900434
